---
title: "Double well diffusion"
output: html_notebook
---

```{r}
library(tidyverse)
library(rstan)
library(loo)
options(mc.cores=4)
rstan_options(auto_write=TRUE)
source("helper.R")
set.seed(1)
```

In this notebook, we consider inference for the following SDE:

\begin{equation}
 dX(t) = \alpha X(t) (\gamma^2 - X(t)^2) dt + \kappa dW(t)
\end{equation}

where $\alpha>0$, $\gamma>0$ and $B>0$. To simulate data from this process, we used an Euler-Maruyama scheme.

```{r}
brownian_motion <- function(T, N) {
  delta_t <- T / N
  dW <- rnorm(N, 0, sqrt(delta_t))
  c(0, cumsum(dW))
}

simulate_double_well <- function(X_0, T, N, alpha, gamma, kappa) {
  W <- brownian_motion(T, N)
  delta_t <- T / N
  X <- vector(length = (N + 1))
  X[1] <- X_0
  for(i in 1:N) {
    X[i + 1] <- X[i] +  alpha * X[i] * (gamma^2 - X[i]^2) * delta_t + kappa * (W[i + 1] - W[i])
  }
  X
}

T <- 10
Npoints <- 5000
X_0 <- 0
alpha <- 2
gamma <- 1
kappa <- 1
X <- simulate_double_well(X_0, T, Npoints, alpha, gamma, kappa)

# observe process less regularly
N <- 500
tfull <- seq(0, T, T / Npoints)
t <- seq(0, T, T / N)
X <- approx(tfull, X, t)$y

# plot
df <- tibble(t, X)
df %>% 
  ggplot(aes(x=t, y=X)) +
  geom_line()
```

Add measurement noise to the data.
```{r}
sigma_n <- 1
Y <- rnorm(length(X), X, sigma_n)
df <- df %>% 
  mutate(Y=Y)
df %>%
  ggplot(aes(x=t, y=Y)) +
  geom_line()
saveRDS(df, "../data/processed/double_well_simulation.rds")

# remove zero time entry as causes issues for ODE integrator in Stan
df <- df %>% 
  filter(t > 0)
```


# Approximate inference
```{r}
model_approx <- stan_model("stan/double_well_approx.stan")

# fit model with many different Zs
init_fn_list <- list(
  alpha=2,
  gamma=1,
  kappa=1,
  sigma_n=1)
Nzs <- c(2, 4, 8, 16, 32, 64, 128)
stan_data <- list(
    nobs=nrow(df),
    t=df$t,
    Y=df$Y,
    X_0=X_0,
    T=T,
    kappa_mean=0,
    kappa_width=10)
lfits <- vector(mode="list", length = length(Nzs))
for(i in seq_along(Nzs)) {
  print(paste0("i = ", i))
  lfits[[i]] <- fit_model_any_z(
    Nzs[i], model_approx, init_fn_list, stan_data,
    iters=4000)
}
```

Looking at estimates of parameters as a function of N
```{r}
extract_param <- function(param_name, fit) {
  rstan::extract(fit, param_name)[[1]]
}
Ns <- c(Nzs)
params <- c("alpha", "gamma", "kappa", "sigma_n")
k <- 1
for(j in seq_along(params)) {
  for(i in seq_along(lfits)) {
    draws <- extract_param(params[j], lfits[[i]])
    temp <- tibble(value=draws,
                   iter=seq_along(draws),
                   N=Ns[i],
                   param=params[j])
    if(k == 1)
      big_df <- temp
    else
      big_df <- big_df %>% bind_rows(temp)
    k <- k + 1
  }
}
big_df <- big_df %>% 
  mutate(param=as.factor(param)) %>% 
  mutate(param=fct_relevel(param, params))
saveRDS(big_df, "../data/processed/double_well_approx_fits.rds")
big_df <- readRDS("../data/processed/double_well_approx_fits.rds") %>% 
  mutate(param=as.character(param)) %>% 
  mutate(param=if_else(param=="sigma_n", "sigma", param)) %>% 
  mutate(param=fct_relevel(param, params))
params <- c("alpha", "gamma", "kappa", "sigma")
short_df <- tribble(
  ~param, ~value,
  "alpha", alpha,
  "gamma", gamma,
  "kappa", kappa,
  "sigma", sigma_n
) %>% 
  mutate(param=as.factor(param)) %>% 
  mutate(param=fct_relevel(param, params))

# remove few outlier values that skew plots
big_df1 <- big_df %>% 
  group_by(param) %>% 
  filter(value <= quantile(value, 0.97))

g <- ggplot(big_df1,
       aes(x=value, colour=as.factor(N))) +
  stat_density(geom="line", position="identity",
               data=big_df1 %>% filter(N != "true") %>%
                 mutate(N=as.numeric(as.character(N)))) +
  scale_color_brewer("N", palette = "Spectral") +
   geom_vline(data=short_df, aes(xintercept=value), linetype=2) +
  facet_wrap(~param, scales="free",
             labeller = label_parsed)
ggsave("../figures/double_well_fits_vs_actual.pdf", g,
       width = 10, height = 6)
```

Extract log-likelihoods to perform model comparison
```{r}
loos <- vector(length = length(lfits), mode="list")
for(i in seq_along(loos)) {
  log_like <- extract_log_lik(lfits[[i]], "loglikelihood")
  loos[[i]] <- loo(log_like)
}
saveRDS(loos, "../data/processed/double_well_loos.rds")
loo_compare(loos)
```

Check convergence
```{r}
rhats <- vector(length=length(lfits))
rhats_higher <- vector(length=length(lfits))
esss <- vector(length=length(lfits))
for(i in seq_along(lfits)) {
  print(i)
  sum_df <- posterior::summarise_draws(lfits[[i]])
  sum_df1 <- sum_df %>% 
    filter(variable%in%c("alpha", "kappa", "gamma", "sigma_n"))
  sum_df2 <- sum_df %>% 
    filter(str_detect(variable, "Z"))
  sum_df <- sum_df1 %>% bind_rows(sum_df2)
  rhats[i] <- sum(sum_df$rhat > 1.01, na.rm = T)
  rhats_higher[i] <- sum(sum_df$rhat > 1.1, na.rm = T)
  bulk_ess <- sum(sum_df$ess_bulk < 400, na.rm = T)
  tail_ess <- sum(sum_df$ess_tail < 400, na.rm = T)
  esss[i] <- bulk_ess + tail_ess
}
conv_df <- tibble(
  rhat=rhats, ess=esss, rhat_1.10=rhats_higher
)
saveRDS(conv_df, "../data/processed/double_well_convergence_diags.rds")
```

# Optimisation for AIC comparison
```{r}
Nzs <- c(2, 4, 8, 16, 32, 64, 128)
lopts <- vector(mode="list", length = length(Nzs))
for(i in seq_along(Nzs)) {
  print(paste0("i = ", i))
  lopts[[i]] <- optimise_model_any_z(
    Nzs[i], model_approx, init_fn_list, stan_data)
}

# plot AIC
N_theta <- 4
lvals <- vector(length = length(Nzs))
for(i in seq_along(lvals))
  lvals[i] <- 2 * Nzs[i] - 2 * lopts[[i]]$value
g <- tibble(N=Nzs, AIC=lvals) %>% 
  ggplot(aes(x=N, y=AIC)) +
  geom_point() +
  geom_line() + 
  scale_x_log10(breaks=Nzs)
ggsave("../figures/double_well_aics.pdf", g,
       width = 8, height = 4)
```