---
title: "Geometric Brownian motion"
output: html_notebook
---

```{r}
library(tidyverse)
library(rstan)
library(loo)
options(mc.cores=4)
rstan_options(auto_write=TRUE)
source("helper.R")
set.seed(1)
```

In this notebook, we perform inference for geometric Brownian motion, which has an SDE:

\begin{equation}
  dX(t) = \theta X(t) + \kappa X(t) dW(t)
\end{equation}

This SDE has a the following probability density, which allows us to exactly simulate from the process:

\begin{equation}
  X(t+\delta t) | X(t) \sim \text{log-normal}(\log X(t) + (\theta - 0.5\kappa^2)\delta_t, \kappa \sqrt{t})
\end{equation}


```{r}
simulate_gbm <- function(X_0, T, N, theta, kappa) {
  delta_t <- T / N
  X <- vector(length = (N + 1))
  X[1] <- X_0
  for(i in 1:N) {
    X[i + 1] <- rlnorm(1, log(X[i]) + (theta - 0.5 * kappa^2) * delta_t, kappa * sqrt(delta_t))
  }
  X
}

T <- 10
N <- 500
X_0 <- 1
theta <- 0.5
kappa <- 0.5
X <- simulate_gbm(X_0, T, N, theta, kappa)
t <- seq(0, T, T/N)

# plot
df <- tibble(t, X)
df %>% 
  ggplot(aes(x=t, y=X)) +
  geom_line()
```

We now add some noise to the data.
```{r}
sigma_n <- 0.1
Y <- rlnorm(length(X), log(X) - 0.5 * sigma_n^2, sigma_n)
df <- df %>% 
  mutate(Y=Y)
df %>% 
  ggplot(aes(x=t, y=Y)) +
  geom_line()
# saveRDS(df, "../data/processed/gbm_simulation.rds")

# remove zero time entry as causes issues for ODE integrator in Stan
df <- df %>% 
  filter(t > 0)
```

# Approximate inference
```{r}
model_approx <- stan_model("stan/gbm_approx.stan")

# fit model with many different Zs
stan_data <- list(
    nobs=nrow(df),
    t=df$t,
    Y=df$Y,
    X_0=X_0,
    T=T,
    kappa_mean=0,
    kappa_width=10)
init_fn_list <- list(
  theta=theta,
  kappa=kappa,
  sigma_n=1)
Nzs <- c(2, 4, 8, 16, 32, 64, 128, 256)
lfits <- vector(mode="list", length = length(Nzs))
for(i in seq_along(Nzs)) {
  print(paste0("i = ", i))
  lfits[[i]] <- fit_model_any_z(
    Nzs[i], model_approx, init_fn_list, stan_data,
    iters=2000)
}
```

Looking at estimates of parameters as a function of N
```{r}
extract_param <- function(param_name, fit) {
  rstan::extract(fit, param_name)[[1]]
}
Ns <- c(Nzs)
params <- c("theta", "kappa", "sigma_n")
k <- 1
for(j in seq_along(params)) {
  for(i in seq_along(lfits)) {
    draws <- extract_param(params[j], lfits[[i]])
    temp <- tibble(value=draws,
                   iter=seq_along(draws),
                   N=Ns[i],
                   param=params[j])
    if(k == 1)
      big_df <- temp
    else
      big_df <- big_df %>% bind_rows(temp)
    k <- k + 1
  }
}
big_df <- big_df %>% 
  mutate(param=as.factor(param)) %>% 
  mutate(param=fct_relevel(param, params))
saveRDS(big_df, "../data/processed/gbm_approx_fits.rds")

params <- c("theta", "kappa", "sigma")
big_df <- readRDS("../data/processed/gbm_approx_fits.rds") %>% 
  mutate(param=as.character(param)) %>% 
  mutate(param=if_else(param=="sigma_n", "sigma", param)) %>% 
  mutate(param=fct_relevel(param, params))

short_df <- tribble(
  ~param, ~value,
  "theta", theta,
  "kappa", kappa,
  "sigma", sigma_n
) %>% 
  mutate(param=as.factor(param)) %>% 
  mutate(param=fct_relevel(param, params))

g <- ggplot(big_df,
       aes(x=value, colour=as.factor(N))) +
  stat_density(geom="line", position="identity",
               data=big_df %>% filter(N != "true") %>%
                 mutate(N=as.numeric(as.character(N)))) +
  scale_color_brewer("N", palette = "Spectral") +
   geom_vline(data=short_df, aes(xintercept=value), linetype=2) +
  facet_wrap(~param, scales="free",
             labeller = label_parsed)
ggsave("../figures/gbm_fits_vs_actual.pdf", g,
       width = 10, height = 6)
```

Plotting the joint distribution of (kappa, sigma) as a function of N
```{r}
g <- big_df %>% 
  pivot_wider(id_cols = c(iter, N),
              values_from = value,
              names_from = param) %>% 
  ggplot(aes(x=kappa, y=sigma, colour=as.factor(N))) +
  geom_point() +
  scale_color_brewer("N", palette = "Spectral") +
  xlab(latex2exp::TeX("$\\kappa$")) +
  ylab(latex2exp::TeX("$\\sigma$"))
ggsave("../figures/gbm_scatter.pdf", g, width = 6,
       height = 4)
```


Extract log-likelihoods to perform model comparison
```{r}
loos <- vector(length = length(lfits), mode="list")
for(i in seq_along(loos)) {
  log_like <- extract_log_lik(lfits[[i]], "loglikelihood")
  loos[[i]] <- loo(log_like)
}
saveRDS(loos, "../data/processed/gbm_loos.rds")
loo_compare(loos)
```

Check convergence
```{r}
rhats <- vector(length=length(lfits))
esss <- vector(length=length(lfits))
for(i in seq_along(lfits)) {
  sum_df <- posterior::summarise_draws(lfits[[i]])
  rhats[i] <- sum(sum_df$rhat > 1.01, na.rm = T)
  bulk_ess <- sum(sum_df$ess_bulk < 400, na.rm = T)
  tail_ess <- sum(sum_df$ess_tail < 400, na.rm = T)
  esss[i] <- bulk_ess + tail_ess
}
conv_df <- tibble(
  rhat=rhats, ess=esss
)
saveRDS(conv_df, "../data/processed/gbm_convergence_diags.rds")
```

# Try quick optimisation then AIC comparison
```{r}
Nzs <- c(2, 4, 8, 16, 32, 64, 128, 256)
lopts <- vector(mode="list", length = length(Nzs))
for(i in seq_along(Nzs)) {
  print(paste0("i = ", i))
  lopts[[i]] <- optimise_model_any_z(
    Nzs[i], model_approx, init_fn_list, stan_data)
}
saveRDS(lopts, "../data/processed/gbm_opts.rds")

# plot AIC
N_theta <- 3
lvals <- vector(length = length(Nzs))
N_theta <- 2
for(i in seq_along(lvals))
  lvals[i] <- 2 * (Nzs[i] + N_theta) - 2 * lopts[[i]]$value
g <- tibble(N=Nzs, AIC=lvals) %>% 
  ggplot(aes(x=N, y=AIC)) +
  geom_point() +
  geom_line() +
  scale_x_log10(breaks=Nzs)
ggsave("../figures/gbm_aics.pdf", g,
       width = 8, height = 4)
```

Plot fit as a function of N
```{r}
for(i in seq_along(lopts)) {
  opt <- lopts[[i]]
  temp_df <- tibble(t=df$t,
                    X=opt$par$X_sim[, 1],
                    N=Nzs[i])
  if(i == 1)
    big_df <- temp_df
  else
    big_df <- big_df %>% bind_rows(temp_df)
}

cols <- brewer.pal(length(Nzs), "Spectral")
cols <- c("#000000", cols)
Ns <- c("actual", Nzs)
big_df <- big_df %>%
  mutate(N=as.character(N)) %>% 
  bind_rows(df %>% mutate(N="actual")) %>% 
  mutate(N=as.factor(N)) %>% 
  mutate(N=fct_relevel(N, Ns))

g1 <- ggplot(df,
       aes(x=t, y=X)) +
  geom_line(data=big_df,
            aes(colour=as.factor(N))) +
  geom_line() +
  scale_color_manual("", values = cols) +
  theme(legend.position = "none")
g1
```
Autocorrelations in errors
```{r}
for(i in seq_along(lopts)) {
  opt <- lopts[[i]]
  temp_df <- tibble(t=df$t,
                    X=opt$par$X_sim[, 1],
                    N=Nzs[i])
  if(i == 1)
    big_df <- temp_df
  else
    big_df <- big_df %>% bind_rows(temp_df)
}

df_true <- df %>% 
  rename(X_true=X) %>% 
  left_join(big_df) %>% 
  mutate(resid=X_true-X)

acf_val <- acf(df_true %>% filter(N==2) %>% pull(resid))
m_acfs <- matrix(nrow = length(acf_val$acf), ncol=length(Nzs))
for(i in seq_along(Nzs)) {
  resids <- df_true %>% filter(N==Nzs[i]) %>% pull(resid)
  acf_val <- acf(resids)$acf[, ,]
  m_acfs[, i] <- acf_val
}
m_acfs <- as.data.frame(m_acfs)
colnames(m_acfs) <- Nzs
m_acfs$lag <- seq_along(m_acfs$`2`)
g2 <- m_acfs %>% 
  pivot_longer(-lag) %>% 
  rename(N=name) %>% 
  mutate(N=as.factor(N)) %>% 
  mutate(N=fct_relevel(N, as.character(Nzs))) %>% 
  ggplot(aes(x=lag, y=value, colour=N)) +
  geom_line() +
  ylab("autocorrelation") +
  scale_colour_brewer("N", palette = "Spectral")
g2
```
```{r}
g <- cowplot::plot_grid(g1, g2, labels = c("A.", "B."))
cowplot::save_plot("../figures/gbm_fit_autocorrelation.pdf", g, base_height = 6, base_width = 12)
```


# AIC scans for different realisations
```{r}
aic_scan <- function(df, Nzs) {
  lopts <- vector(mode="list", length = length(Nzs))
  stan_data <- list(
    nobs=nrow(df),
    t=df$t,
    Y=df$Y,
    X_0=X_0,
    T=T,
    kappa_mean=0,
    kappa_width=10)
  
  for(i in seq_along(Nzs)) {
    print(paste0("i = ", i))
    lopts[[i]] <- optimise_model_any_z(
      Nzs[i], model_approx, init_fn_list, stan_data)
  }
  lvals <- vector(length = length(Nzs))
for(i in seq_along(lvals))
  lvals[i] <- 2 * (Nzs[i] + N_theta) - 2 * lopts[[i]]$value
  lvals
}

simulate_dataset <- function(theta=0.5, kappa=0.5, sigma_n=0.1) {
  T <- 10
  N <- 500
  X_0 <- 1
  X <- simulate_gbm(X_0, T, N, theta, kappa)
  t <- seq(0, T, T/N)
  df <- tibble(t, X)
  Y <- rlnorm(length(X), log(X) - 0.5 * sigma_n^2, sigma_n)
  df <- df %>% 
    mutate(Y=Y)
  df %>% 
    filter(t > 0)
}

Nzs <- c(2, 4, 8, 16, 32, 64)
Nrepeats <- 5
m_aics <- matrix(nrow = length(Nzs),
                 ncol = Nrepeats)
for(i in 1:Nrepeats) {
  print(paste0("iter = ", i))
  df_temp <- simulate_dataset()
  m_aics[, i] <- aic_scan(df_temp, Nzs)
}

m_aics <- as.data.frame(m_aics)
colnames(m_aics) <- 1:Nrepeats
m_aics <- m_aics %>% 
  mutate(N=Nzs)
saveRDS(m_aics, "../data/processed/gbm_aics_different_samples.rds")
m_aics <- readRDS("../data/processed/gbm_aics_different_samples.rds")
g <- m_aics %>%
  pivot_longer(-N) %>%
  mutate(name=as.factor(name)) %>%
  ggplot(aes(x=N, y=value, colour=name)) +
  geom_line() +
  geom_point() +
  scale_x_log10(breaks=Nzs) +
  scale_color_brewer(palette = "Dark2") +
  theme(legend.position = "none") +
  ylab("AIC")
g
ggsave("../figures/gbm_aics_repeats.pdf", g,
       width = 8, height = 4)
apply(m_aics, 2, which.min)
```

# Perform AIC scan for completely different parameter set
```{r}
theta <- -0.5
kappa <- 0.1
sigma_n <- 0.2

Nzs <- c(2, 4, 8, 16, 32, 64)
df_temp <- simulate_dataset(theta=theta,
                            kappa=kappa,
                            sigma_n=sigma_n)
df_temp %>% 
  ggplot(aes(x=t, y=X)) +
  geom_line()
lvals <- aic_scan(df_temp, Nzs)
g <- tibble(N=Nzs, AIC=lvals) %>% 
  ggplot(aes(x=N, y=AIC)) +
  geom_point() +
  geom_line() +
  scale_x_log10(breaks=Nzs)
ggsave("../figures/gbm_aics_different_parameters.pdf", g,
       width = 8, height = 4)
```


