---
title: "Geometric Brownian motion"
output: html_notebook
---

```{r}
library(tidyverse)
library(rstan)
options(mc.cores=4)
rstan_options(auto_write=TRUE)
source("helper.R")
```

In this notebook, we perform inference for geometric Brownian motion, which has an SDE:

\begin{equation}
  dX(t) = \theta X(t) + \kappa X(t) dW(t)
\end{equation}

This SDE has a the following probability density, which allows us to exactly simulate from the process:

\begin{equation}
  X(t+\delta t) | X(t) \sim \text{log-normal}(\log X(t) + (\theta - 0.5\kappa^2)\delta_t, \kappa \sqrt{t})
\end{equation}


```{r}
simulate_gbm <- function(X_0, T, N, theta, kappa) {
  delta_t <- T / N
  X <- vector(length = (N + 1))
  X[1] <- X_0
  for(i in 1:N) {
    X[i + 1] <- rlnorm(1, log(X[i]) + (theta - 0.5 * kappa^2) * delta_t, kappa * sqrt(delta_t))
  }
  X
}

T <- 10
N <- 500
X_0 <- 1
theta <- 0.5
kappa <- 0.5
X <- simulate_gbm(X_0, T, N, theta, kappa)
t <- seq(0, T, T/N)

# plot
df <- tibble(t, X)
df %>% 
  ggplot(aes(x=t, y=X)) +
  geom_line()
```

We now add some noise to the data.
```{r}
sigma_n <- 0.1
Y <- rlnorm(length(X), log(X) - 0.5 * sigma_n^2, sigma_n)
df <- df %>% 
  mutate(Y=Y)
df %>% 
  ggplot(aes(x=t, y=Y)) +
  geom_line()
saveRDS(df, "../data/processed/gbm_simulation.rds")
```

# Approximate inference
```{r}
model_approx <- stan_model("stan/gbm_approx.stan")

# fit model with many different Zs
init_fn <- function() {
    list(theta=theta,
         Z=as.array(rep(1, N_zs)),
         sigma=sigma,
         sigma_n=1)
  }
Nzs <- c(2, 4, 8, 16, 32, 64, 128)
lfits <- vector(mode="list", length = length(Nzs))
for(i in seq_along(Nzs)) {
  print(paste0("i = ", i))
  lfits[[i]] <- fit_model_any_z(NZs[i], model_approx,
                                init_fn)
}
```

Looking at estimates of parameters as a function of N
```{r}
extract_param <- function(param_name, fit) {
  rstan::extract(fit, param_name)[[1]]
}
Ns <- c("true", Nzs)
params <- c("theta", "kappa", "sigma_n")
k <- 1
for(j in seq_along(params)) {
  for(i in seq_along(lfits)) {
    draws <- extract_param(params[j], lfits[[i]])
    temp <- tibble(value=draws,
                   iter=seq_along(draws),
                   N=Ns[i],
                   param=params[j])
    if(k == 1)
      big_df <- temp
    else
      big_df <- big_df %>% bind_rows(temp)
    k <- k + 1
  }
}
big_df <- big_df %>% 
  mutate(param=as.factor(param)) %>% 
  mutate(param=fct_relevel(param, params))
saveRDS(big_df, "../data/processed/gbm_approx_fits.rds")

short_df <- tribble(
  ~param, ~value,
  "theta", theta,
  "sigma", sigma,
  "sigma_n", sigma_n
) %>% 
  mutate(param=as.factor(param)) %>% 
  mutate(param=fct_relevel(param, params))

g <- ggplot(big_df,
       aes(x=value)) +
  geom_density(aes(colour=as.factor(N)), data=big_df %>% filter(N != "true") %>%
                 mutate(N=as.numeric(as.character(N)))) +
  geom_density(aes(colour=as.factor(N)), data=big_df %>% filter(N == "true"),
               colour="black", linetype=2) +
  scale_color_brewer("N", palette = "Spectral") +
   geom_vline(data=short_df, aes(xintercept=value), linetype=3) +
  facet_wrap(~param, scales="free")
ggsave("../figures/gbm_fits_vs_actual.pdf", g,
       width = 10, height = 6)
```

Extract log-likelihoods to perform model comparison
```{r}
loos <- vector(length = length(lfits))
for(i in seq_along(loos)) {
  log_like <- extract_log_lik(lfits[[i]], "loglikelihood")
  loos[[i]] <- loo(log_like)
}
saveRDS(loos, "../data/processed/gbm_loos.rds")
loo_compare(loos)
```