---
title: "KL expansion of Brownian motion"
output: html_notebook
---

```{r}
library(tidyverse)
set.seed(1)
```

The below functions are used to generate an approximate realisation of a standard Brownian motion through its KL expansion, which also returns the $Z_i$s used to generate it:

\begin{equation}
W(t) \approx \sqrt{2T}\sum_{i=1}^N \frac{2}{(2i - 1)\pi} Z_i \sin\left(\frac{(2i - 1)\pi t}{2T} \right).
\end{equation}

We assume that the process is observed at $t=0, \delta t, 2\delta t,...,T$.

```{r}
int_phi <- function(t, i, T) {
  sqrt(2 * T) * 2 / ((2 * i - 1) * pi) * sin((2 * i - 1) * pi * t / (2 * T))
}

W_simulate <- function(Zs, ts, T) {
  one_to_N <- 1:length(Zs)
  W <- vector(length = length(ts))
  for(i in seq_along(ts))
    W[i] <- sum(Zs * int_phi(ts[i], one_to_N, T))
  W
}

simulate_brownian_kl_expansion <- function(T, delta_t, N) {
  Zs <- rnorm(N, 0, 1)
  ts <- seq(0, T, delta_t)
  W <- W_simulate(Zs, ts, T)
  list(t=ts, W=W, Zs=Zs)
}
```

Let's test them out for $T=10$ and $\delta t = 0.01$ with $N=10,000$.
```{r}
N <- 10000
T <- 10
delta_t <- 0.01
res <- simulate_brownian_kl_expansion(T, delta_t, N)
ts <- res$t
W <- res$W
Zs <- res$Z

# plot Brownian  motion realisation
df <- tibble(t=ts, W=W)
g1 <- df %>% 
  ggplot(aes(x=t, y=W)) +
  geom_line()
g1
```

Let's now progressively approximate the Brownian motion as we increase $N$. To do so, we use the Zs we randomly sampled above to generate the specific Brownian motion path.
```{r}
Ns <- c(1, 2, 4, 8, 16, 32, 64, 128, 256, 512, 1024)
Zs <- res$Zs
for(i in seq_along(Ns)) {
  N <- Ns[i]
  Zs_used <- Zs[1:N]
  W <- W_simulate(Zs_used, ts, T)
  df_tmp <- tibble(t=ts, W=W, N=N)
  # stack tibbles for plotting
  if(i == 1)
    df_sims <- df_tmp
  else
    df_sims <- df_sims %>% bind_rows(df_tmp)
}
df_sims <- df_sims %>% 
  mutate(type="approximations")
df <- df %>% 
  mutate(type="actual")

# plot
g2 <- ggplot(df_sims, aes(x=t, y=W, group=N)) +
  geom_line(aes(colour=as.factor(N))) +
  geom_line(data=df) +
  scale_color_viridis_d("N")
g2

# ggsave("../figures/brownian_approximation_kl.pdf", g2, width = 8, height = 4)
```

It's actually quite hard to see the approximations in the above, so let's zoom in. This shows that as $N$ increases, we obtain a better approximation of the particular Brownian path. The lower frequency components represent the broader changes in the path; the higher frequency components dictate variation over shorter timescales.

```{r}
ggplot(df_sims %>% filter(t < 2.5),
       aes(x=t, y=W, group=N)) +
  geom_line(data=df %>% filter(t < 2.5)) +
  geom_line(aes(colour=N)) +
  scale_color_viridis_c()
```
In the next figure, we decompose a (very approximate) Brownian path (drawn using $n=10$) into its constituent functions. Whilst these draws are random, these typically show that the longest wavelength components (i.e. low frequency) components dominate giving a Brownian motion process its random walk type behaviour.
```{r}
int_phi <- function(t, i, T) {
  sqrt(2 * T) * 2 / ((2 * i - 1) * pi) * sin((2 * i - 1) * pi * t / (2 * T))
}
N <- 10
T <- 10
delta_t <- 0.01
res <- simulate_brownian_kl_expansion(T, delta_t, N)
ts <- res$t
Zs <- res$Zs
W_parts <- matrix(nrow = length(ts),
                  ncol = N)
for(i in 1:N) {
  W_parts[, i] <- Zs[i] * int_phi(ts, i, T)
}
W_parts <- as_tibble(W_parts) %>% 
  mutate(t=ts) %>% 
  pivot_longer(-t) %>% 
  rename(component=name) %>% 
  mutate(component=gsub("V", "", component)) %>% 
  mutate(component=as.numeric(as.character(component)))

# plot Brownian path
tibble(t=ts, W=res$W) %>% 
  ggplot(aes(x=t, y=W)) +
  geom_line()

# plot components
W_parts %>% 
  ggplot(aes(x=t, y=value,
             group=component,
             colour=as.factor(component))) +
  geom_line() +
  scale_color_brewer("Component",
                     palette = "Spectral") +
  ylab("value")
```

We now show that white noise can be approximated via a KL expansion.
```{r}
phi <- function(t, i, T) {
  sqrt(2 / T) * cos((2 * i - 1) * pi * t / (2 * T))
}

dW_simulate <- function(Zs, ts, T) {
  one_to_N <- 1:length(Zs)
  dW <- vector(length = length(ts))
  for(i in seq_along(ts))
    dW[i] <- sum(Zs * phi(ts[i], one_to_N, T))
  dW
}

simulate_white_noise_kl_expansion <- function(T, delta_t, N) {
  Zs <- rnorm(N, 0, 1)
  ts <- seq(0, T, delta_t)
  dW <- dW_simulate(Zs, ts, T)
  list(t=ts, dW=dW, Zs=Zs)
}

T <- 5
res <- simulate_white_noise_kl_expansion(T, delta_t, 10000)
ts <- res$t
dW <- res$dW
Zs <- res$Z
df <- tibble(t=ts, dW=dW)

# reconstruct it with ever increasing N
Ns <- 2^seq(0, 13, 1)
Zs <- res$Zs
for(i in seq_along(Ns)) {
  N <- Ns[i]
  Zs_used <- Zs[1:N]
  dW <- dW_simulate(Zs_used, ts, T)
  df_tmp <- tibble(t=ts, dW=dW, N=N)
  # stack tibbles for plotting
  if(i == 1)
    df_sims <- df_tmp
  else
    df_sims <- df_sims %>% bind_rows(df_tmp)
}
df_sims <- df_sims %>% 
  mutate(type="approximations")
df <- df %>% 
  mutate(type="actual")

# plot
df_sims_fewer <- df_sims %>% 
  filter(N %in% c(1, 16, 256, 4096))
g1 <- ggplot(df_sims_fewer, aes(x=t, y=dW, group=N)) +
  geom_line(aes(colour=as.factor(N))) +
  geom_line(data=df) +
  scale_color_viridis_d("N") +
  theme(legend.position = c(0.75, 0.9)) +
  xlim(0, 2) +
  guides(colour=guide_legend(nrow=2,byrow=TRUE))
g1

# Look at correlation between dW and approximation as N increases
df_both <- df_sims %>% 
  select(-type) %>% 
  left_join(df %>% rename(true=dW) %>%
              select(-type)) %>% 
  rename(approx=dW)

# plot correlation
g2 <- df_both %>% 
  group_by(N) %>% 
  summarise(rho=cor(true, approx)) %>% 
  ggplot(aes(x=N, y=rho)) +
  geom_point() +
  scale_x_log10() +
  ylab(latex2exp::TeX("$\\rho$"))

g <- cowplot::plot_grid(g1, g2, nrow=1, labels=c("A.", "B."))
cowplot::save_plot("../figures/white_noise_kl_approximation.pdf", g, base_width = 12, base_height = 6)
g
```

