---
title: "Square root process"
output: html_notebook
---

```{r}
library(tidyverse)
library(rstan)
library(loo)
library(RColorBrewer)
options(mc.cores=4)
rstan_options(auto_write=TRUE)
source("helper.R")
set.seed(1)
```

In this notebook, we perform inference for the Orstein-Uhlenbeck (OU) process, which is described by the following SDE:

\begin{equation}
  dX(t) = dt + \kappa \sqrt{X(t)} dW(t).
\end{equation}

Throughout, we assume that $X(0)=10$ which is known and $\kappa=2$, which is not. The observations are corrupted by Gaussian measurement noise of the form:

\begin{equation}
  Y(t) \stackrel{i.i.d.}{\sim} \text{normal}(X(t), \sigma).
\end{equation}

We first simulate some data from the OU process with measurement noise using known parameters. Exact simulation of this process is possible, since there is a closed-form solution to this process:

\begin{equation}
X(t) = (W(t) + \sqrt{X(0)})^2
\end{equation}

```{r}
simulate_sqrt <- function(X_0, T, N) {
  W <- brownian_motion(T, N)
  X <- (W + X_0)^2
  X
}

T <- 10
N <- 500
X_0 <- 0
X <- simulate_sqrt(X_0, T, N)
t <- seq(0, T, T/N)

# plot
df <- tibble(t, X)
df %>% 
  ggplot(aes(x=t, y=X)) +
  geom_line()
```
Now add Gaussian measurement noise
```{r}
sigma_n <- 1
Y <- rnorm(length(X), X, sigma_n)
df <- df %>% 
  mutate(Y=Y)
df %>%
  ggplot(aes(x=t, y=Y)) +
  geom_line()
saveRDS(df, "../data/processed/sqrt_simulation.rds")

# remove zero time entry as causes issues for ODE integrator in Stan
df <- df %>% 
  filter(t > 0)
```

# Optimisation for AIC comparison
```{r}
model_approx <- stan_model("stan/sqrt_approx.stan")
# expose_stan_functions(model_approx)

params <- c(2, 10, -1, 2)
X_sim <- int_x(500, 0.0, df$t, params, c(1, 2), c(2))
plot(unlist(X_sim))

# fit model with many different Zs
init_fn_list <- list(
  kappa=2,
  sigma_n=1)
Nzs <- c(2, 4, 8, 16, 32, 64, 128)
stan_data <- list(
    nobs=nrow(df),
    t=df$t,
    Y=df$Y,
    X_0=X_0,
    T=T,
    kappa_mean=0,
    kappa_width=10)

Nzs <- c(2, 4, 8, 16, 32)
lopts <- vector(mode="list", length = length(Nzs))
for(i in seq_along(Nzs)) {
  print(paste0("i = ", i))
  lopts[[i]] <- optimise_model_any_z(
    Nzs[i], model_approx, init_fn_list, stan_data)
}

# plot AIC
N_theta <- 3 # theta, kappa, sigma
lvals <- vector(length = length(Nzs))
for(i in seq_along(lvals))
  lvals[i] <- 2 * (N_theta + Nzs[i]) - 2 * lopts[[i]]$value
g <- tibble(N=Nzs, AIC=lvals) %>% 
  ggplot(aes(x=N, y=AIC)) +
  geom_point() +
  geom_line() +
  scale_x_log10(breaks=Nzs)
ggsave("../figures/sqrt_aics.pdf", g,
       width = 8, height = 4)
```
