---
title: "Wiener velocity"
output: html_notebook
---

```{r}
library(tidyverse)
library(rstan)
options(mc.cores=4)
rstan_options(auto_write=TRUE)
source("helper.R")
library(mvtnorm)
```


In this notebook, we perform inference on noisy measurements of the following bivariate SDE:

\begin{equation}
    dX(t) = \begin{bmatrix}
   0 & 1\\
   0 & 0 
   \end{bmatrix} X(t) dt + \begin{bmatrix}
   0\\
   \sqrt{q} 
   \end{bmatrix}dW(t)
\end{equation}

and where $X(t) = [X_1(t), X_2(t)]'$.

Exact simulation for this model is possible, since the transition density is multivariate normally distributed (eq. 17: https://advancesindifferenceequations.springeropen.com/articles/10.1186/s13662-019-2214-1). The mean of the process is given by:

\begin{equation}
mu(t + \delta t) = \exp(-\theta \delta t) X[t] = \begin{bmatrix}1 & \delta t \\
 0 & 1 \\
 \end{bmatrix} X[t]
\end{equation}

where $\theta=\begin{bmatrix} 0 & -1\\ 0 & 0\\ \end{bmatrix}$. The covariance matrix is given by $\Sigma = \begin{bmatrix}\frac{q \delta t^3}{3} & \frac{q \delta t^2}{2} \\ \frac{q \delta t^2}{2} & q \delta t \\ \end{bmatrix}$.

```{r}
simulate_wiener <- function(X_1_0, X_2_0, T, N, q) {
  delta_t <- T / N
  X <- matrix(nrow=(N + 1), ncol=2)
  M <- matrix(c(1, delta_t, 0, 1), byrow = TRUE, ncol = 2)
  Sigma <- matrix(c((q * delta_t^3) / 3, (q * delta_t^2) / 2, (q * delta_t^2) / 2, q * delta_t), byrow = TRUE, ncol = 2)
  X[1, ] <- c(X_1_0, X_2_0)
  for(i in 1:N) {
    X[i + 1, ] <- rmvnorm(1, M %*% X[i, ], Sigma)
  }
  X
}

T <- 10
N <- 500
X_1_0 <- 1
X_2_0 <- 0
q <- 1
X <- simulate_wiener(X_1_0, X_2_0, T, N, q)
t <- seq(0, T, T/N)
df <- tibble(t=t, v=X[, 1], a=X[, 2])
df %>% 
  pivot_longer(-t) %>% 
  ggplot(aes(x=t, y=value, colour=name)) +
  geom_line() +
  scale_color_brewer("Series", palette = "Dark2")
```
Now adding Gaussian measurement noise to the observed velocity series.
```{r}
sigma_n <- 1
y <- rnorm(nrow(X), X[, 1], sigma_n)
df <- df %>% 
  mutate(Y=y)
df %>% 
  pivot_longer(-t) %>% 
  ggplot(aes(x=t, y=value, colour=name)) +
  geom_line() +
  scale_color_brewer("Series", palette = "Dark2")
```

# Exact inference
Since the SDE is linear, it is possible to use Kalman filtering to calculate the likelihood, and we can perform exact inference.
```{r}
model <- stan_model("stan/wiener_velocity.stan")
stan_data <- list(
  nobs=nrow(df),
  t=df$t,
  Y=df$Y,
  X_1_0=X_1_0,
  X_2_0=X_2_0,
  delta_t = T / nobs
)
fit <- sampling(model, data=stan_data, iter=2000,
                chains=4)
saveRDS(fit, "../data/processed/wiener_velocity_exact_fit.rds")
```

