---
title: "Wiener velocity"
output: html_notebook
---

```{r}
library(tidyverse)
library(rstan)
library(mvtnorm)
library(RColorBrewer)
options(mc.cores=4)
rstan_options(auto_write=TRUE)
source("helper.R")
set.seed(1)
```


In this notebook, we perform inference on noisy measurements of the following bivariate SDE:

\begin{equation}
    dX(t) = \begin{bmatrix}
   0 & 1\\
   0 & 0 
   \end{bmatrix} X(t) dt + \begin{bmatrix}
   0\\
   \sqrt{q} 
   \end{bmatrix}dW(t)
\end{equation}

and where $X(t) = [X_1(t), X_2(t)]'$.

Exact simulation for this model is possible, since the transition density is multivariate normally distributed (eq. 17: https://advancesindifferenceequations.springeropen.com/articles/10.1186/s13662-019-2214-1). The mean of the process is given by:

\begin{equation}
mu(t + \delta t) = \exp(-\theta \delta t) X[t] = \begin{bmatrix}1 & \delta t \\
 0 & 1 \\
 \end{bmatrix} X[t]
\end{equation}

where $\theta=\begin{bmatrix} 0 & -1\\ 0 & 0\\ \end{bmatrix}$. The covariance matrix is given by $\Sigma = \begin{bmatrix}\frac{q \delta t^3}{3} & \frac{q \delta t^2}{2} \\ \frac{q \delta t^2}{2} & q \delta t \\ \end{bmatrix}$.

```{r}
simulate_wiener <- function(X_1_0, X_2_0, T, N, q) {
  delta_t <- T / N
  X <- matrix(nrow=(N + 1), ncol=2)
  M <- matrix(c(1, delta_t, 0, 1), byrow = TRUE, ncol = 2)
  Sigma <- matrix(c((q * delta_t^3) / 3, (q * delta_t^2) / 2, (q * delta_t^2) / 2, q * delta_t), byrow = TRUE, ncol = 2)
  X[1, ] <- c(X_1_0, X_2_0)
  for(i in 1:N) {
    X[i + 1, ] <- rmvnorm(1, M %*% X[i, ], Sigma)
  }
  X
}

T <- 10
N <- 500
X_1_0 <- 1
X_2_0 <- 0
q <- 1
X <- simulate_wiener(X_1_0, X_2_0, T, N, q)
t <- seq(0, T, T/N)
df <- tibble(t=t, v=X[, 1], a=X[, 2])
df %>% 
  pivot_longer(-t) %>% 
  ggplot(aes(x=t, y=value, colour=name)) +
  geom_line() +
  scale_color_brewer("Series", palette = "Dark2")
```
Now adding Gaussian measurement noise to the observed velocity series.
```{r}
sigma_n <- 1
y <- rnorm(nrow(X), X[, 1], sigma_n)
df <- df %>% 
  mutate(Y=y)
df %>% 
  pivot_longer(-t) %>% 
  ggplot(aes(x=t, y=value, colour=name)) +
  geom_line() +
  scale_color_brewer("Series", palette = "Dark2")
saveRDS(df, "../data/processed/wiener_simulation.rds")

# remove zero time entry as causes issues for ODE integrator in Stan
df <- df %>% 
  filter(t > 0)
```

# Exact inference
Since the SDE is linear, it is possible to use Kalman filtering to calculate the likelihood, and we can perform exact inference.
```{r}
model <- stan_model("stan/wiener_velocity.stan")
stan_data <- list(
  nobs=nrow(df),
  t=df$t,
  Y=df$Y,
  X_1_0=X_1_0,
  X_2_0=X_2_0,
  delta_t = T / N
)
fit <- sampling(model, data=stan_data, iter=2000,
                chains=4)
saveRDS(fit, "../data/processed/wiener_velocity_exact_fit.rds")
fit <- readRDS("../data/processed/wiener_velocity_exact_fit.rds")
```

# Approximate inference
```{r}
model_approx <- stan_model("stan/wiener_velocity_approx.stan")
stan_data <- list(
    nobs=nrow(df),
    t=df$t,
    Y=df$Y,
    X_1_0=X_1_0,
    X_2_0=X_2_0,
    delta_t = T / N,
    N=2,
    T=T
)
init_fn_list <- list(
  q=1, sigma_n=1)
Nzs <- c(2, 4, 8, 16, 32, 64)
lfits <- vector(mode="list", length = length(Nzs))
for(i in seq_along(Nzs)) {
  print(paste0("i = ", i))
  lfits[[i]] <- fit_model_any_z(
    Nzs[i], model_approx, init_fn_list, stan_data,
    ndims=1, iters=2000)
}
```

Looking at estimates of parameters as a function of N
```{r}
extract_param <- function(param_name, fit) {
  rstan::extract(fit, param_name)[[1]]
}
Ns <- c("true", Nzs)
params <- c("q", "sigma_n")
lfits <- c(fit, lfits)
k <- 1
for(j in seq_along(params)) {
  for(i in seq_along(lfits)) {
    draws <- extract_param(params[j], lfits[[i]])
    temp <- tibble(value=draws,
                   iter=seq_along(draws),
                   N=Ns[i],
                   param=params[j])
    if(k == 1)
      big_df <- temp
    else
      big_df <- big_df %>% bind_rows(temp)
    k <- k + 1
  }
}
big_df <- big_df %>% 
  mutate(param=as.factor(param)) %>% 
  mutate(param=fct_relevel(param, params))
saveRDS(big_df, "../data/processed/wiener_approx_fits.rds")
Ns <- c(rev(Ns[-1]), "actual")
big_df <- readRDS("../data/processed/wiener_approx_fits.rds") %>% 
  mutate(param=as.character(param)) %>% 
  mutate(param=if_else(param=="sigma_n", "sigma", param)) %>% 
  mutate(param=fct_relevel(param, params)) %>% 
  mutate(N=if_else(N=="true", "actual", N)) %>% 
  mutate(N=fct_relevel(N, Ns)) %>% 
  mutate(type=if_else(N=="actual", "actual", "approx")) %>% 
  mutate(type=fct_relevel(type, "approx", "actual"))

cols <- brewer.pal(length(Ns) - 1, "Spectral")
cols <- c(cols, "#000000")

g <- ggplot(big_df,
       aes(x=value, colour=N)) +
  stat_density(geom="line", position="identity",
               aes(linetype=N)) +
  scale_color_manual("", values = cols) +
  scale_linetype_manual("", values=c(rep(1, length(Ns) - 1), 2)) +
  facet_wrap(~param, scales="free",
             labeller = label_parsed) +
  guides(colour = guide_legend(reverse=T),
         linetype = guide_legend(reverse=T))
ggsave("../figures/wiener_fits_vs_actual.pdf", g,
       width = 10, height = 6)
```

Extract log-likelihoods to perform model comparison
```{r}
lfits <- lfits[-1]
loos <- vector(length = length(lfits), mode="list")
for(i in seq_along(loos)) {
  log_like <- loo::extract_log_lik(lfits[[i]], "loglikelihood")
  loos[[i]] <- loo(log_like)
}
saveRDS(loos, "../data/processed/wiener_loos.rds")
loo_compare(loos)
```

# Optimisation for AIC comparison
```{r}
Nzs <- c(2, 4, 8, 16, 32, 64)
lopts <- vector(mode="list", length = length(Nzs))
for(i in seq_along(Nzs)) {
  print(paste0("i = ", i))
  lopts[[i]] <- optimise_model_any_z(
    Nzs[i], model_approx, init_fn_list, stan_data,
    ndims = 1)
}

# plot AIC
N_theta <- 2
lvals <- vector(length = length(Nzs))
for(i in seq_along(lvals))
  lvals[i] <- 2 * (Nzs[i] + N_theta) - 2 * lopts[[i]]$value
g <- tibble(N=Nzs, AIC=lvals) %>%
  filter(N > 2) %>% 
  ggplot(aes(x=N, y=AIC)) +
  geom_point() +
  geom_line() +
  scale_x_log10()
ggsave("../figures/wiener_aics.pdf", g,
       width = 8, height = 4)
```
